### 2018-10-02 paper/paper.tex
Edits by Stu

% Please keep the abstract below 300 words
\section*{Abstract}
When it comes to folding, the RNA energy landscape is particularly rugged, densely populated with 
energy-favorable pairings of nucleotide sequences, most of which are not a part of the molecule's native structure or structures. Yet many non-coding RNA molecules spontaneously fold to a native structure when placed in a suitable chemical environment at an appropriate temperature.
Levinthal's convincing arguments\cite{Levinthal1969-hi} (aka Levinthal's paradox) about proteins apply to RNA molecules as well, perhaps even more compellingly: arrival to thermal equilibrium should take a great deal more time than the biological lifetime of the organism. Evidently, mechanisms have evolved that contribute to a directed and efficient folding process, and quite possibly the active configuration is not at all like a sample from an equilibrium distribution, but rather a metastable state, provided only that it will typically last long enough for its purpose. 
Various measures of the difficulty in reaching a stable configuration, whether or not it is a true equilibrium,
are proposed based on the number of possible Watson-Crick or G$\cdot$U wobble pairings at each location. We call these measures ``ambiguity indexes,'' and focus on two, each of which is a function of both the primary and any given proposed secondary structure. Comparing families of RNA molecules that appear in ribonucleoproteins (``bound'' molecules) to RNA molecules that do not operate as nucleoproteins (``unbound'' molecules), we find that the unbound molecules have systematically lower ambiguity indexes, {\em provided that the indexes are computed using the comparative-analysis \cite{Gutell1992-hu} structure} (also known as the ``gold standard''). The effect largely disappears when the indexes are based, instead, on the so-called minimum-free energy (MFE) structures \cite{Mathews1999-hc,Zuker1999-rc}. Since MFE structures are derived from an approximation of the equilibrium distribution, we interpret these and other related results as evidence for metastable but not necessarily thermal equilibrium secondary structure.

@@

Because of the time-consuming nature of experimental determination of RNA structures, a considerable amount of work has been put into computational (as opposed to experimental) approaches. For secondary structure prediction, comparative analysis is the gold standard  \cite{Gutell1992-hu}. But accurate comparative analysis typically requires a large number of homologous sequences, which may or may not be available. On the other hand, the prevailing {\em computational} approaches are based on an important and controversial assumption\cite{Flamm2008-ir}. Namely, that the structures of  non-coding RNA molecules, {\em in vivo}, are in thermal equilibrium. If this were the case, and if the configuration energies could be accurately specified, then secondary (and in principle, even tertiary) structures could be explored through Monte Carlo sampling from the Boltzmann equilibrium distribution. Alternatively, depending on the nature of the energy landscape, secondary structures could be approximated by identifying the minimum free energy (MFE) configuration  \cite{Mathews1999-hc,Zuker1999-rc}. However, even if we put aside the necessity for approximations, the biological relevance of equilibrium configurations has been a source of misgivings at least since 1969, when Levinthal pointed out that the time required to equilibrate might be too long by many orders of magnitude  \cite{Levinthal1969-hi}. In light of these observations, and considering the ``frustrated'' nature of the folding landscape, many have argued that when it comes to structure prediction for macromolecules, kinetic accessibility is more relevant than equilibrium thermodynamics  \cite{Higgs2000-gj,Flamm2008-ir,Baker1994-px}.

The primary structures of RNA molecules typically afford many opportunities to form short or medium-length stems\footnote{
By which we will mean sequences of G$\cdot$U (``wobble pairs'') and/or Watson-Crick pairs.},
most of which do not participate in the native structure. This situation not only makes it hard for the biologist to accurately predict secondary structure, but equally challenges the molecule to avoid these ``energetic traps.'' Once formed, they require a large amount of energy (not to mention time) to be unformed. By what mechanisms might these traps be discouraged, if not actually avoided? 
Ideally, the only stems available in the primary structure would be those that participate in the secondary structure, but this is obviously too restrictive for all but the shortest or simplest of the structural RNA molecules. Still, there can be little doubt that the necessity for repeatable and efficient folding produces a selective pressure against the more disruptive ambiguities---e.g.
a sequence that can form half of either of two stems, both of which are kinetically accessible and stable, but only one of which is native, or a sequence that is unpaired in the native structure but can nonetheless form a stem that is stereochemically inconsistent with the native structure. By this reasoning, which emphasizes kinetics rather than equilibrium, {\em per se}, we might expect to find information in the intra-molecular ambiguities about the folding process, or even the native structure.

We will introduce quantitative measures of ambiguity and demonstrate their statistical relationships to native secondary structure and to qualitative distinctions between RNA families.
Using a somewhat arbitrary convention, we will
refer to sequences of length four (four consecutive nucleotides) as {\em segments}. Keeping this convention in mind, each line of statistical
evidence is based on the same formal definition of the ``local ambiguity'' of a given segment. This is simply the number of its complementary pairs in the molecule. 
When we refer to the {\em location} of a segment, we will mean the location of the first element of the segment, counting from 5$^\prime$ to 3$^\prime$, and  
when we refer to the local ambiguity of a location, we will mean the local ambiguity of the segment at that location. 
Local ambiguity is an intrinsic property of the primary structure. We will be interested in exposing its statistical relationships to any given candidate secondary structure of the same molecule. For this purpose, for any particular secondary structure we distinguish three different kinds of locations:
\begin{description}
	\item[\textit{Single}:] Locations where all nucleotides in the corresponding segment are unpaired in the secondary structure;
	\item[\textit{Double}:] Locations where all nucleotides in the corresponding segment are paired in the secondary structure;
	\item[\textit{Transitional}:] Locations where some nucleotides in the corresponding segment are paired and others are unpaired in the secondary structure.
\end{description}
{\em Double} and {\em transitional} locations participate in the candidate secondary structure, while {\em single} locations do not.
As a general rule, local ambiguities measured at any of these locations will increase with the length of the molecule---there are simply more opportunities to find complementary sequences.
Instead of using the local ambiguities themselves, we focus on the {\em differences} between ambiguities in and around stems ({\em double} and {\em transitional} locations) from those at unpaired ({\em single}) locations. In particular, we defined the ``T-S ambiguity index'' to be the difference between the average local ambiguity at {\em transitional} locations minus the average at {\em single} locations. We defined the ``D-S ambiguity index'' similarly, but adjusted for an evident bias at {\em double} locations, most of which will have at least one complementary pair somewhere in the molecule. 
(The exceptions being bulges, which account for about 20\% of the {\em double} locations in our data set.)
With this in mind, we defined the D-S ambiguity index to be the average local ambiguity at those {\em double} locations that have at least one ambiguity minus the average over those {\em single} locations that have at least one ambiguity\footnote{
An argument could be made for including bulges in the definition of the {\em local ambiguity} in the first place. We have not explored this or other variations, though there is not shortage of arguably useful descriptive statistics for capturing the notion of ambiguity.}.

@@

How are
we to gauge $d_\text{T-S}$ and $d_\text{D-S}$, and how are we to compare their values across different RNA families? 
Consider the following experiment: for a given RNA molecule we create a ``surrogate'' which has the same nucleotides, and in fact the same counts of {\em all} four-tuple segments as the original molecule, but is otherwise ordered randomly. If ACCU appeared eight times in the original molecule, then it appears eight times in the surrogate, and the same can be said of all sequences of four successive nucleotides---the frequency of each of the $4^4$ possible segments is preserved in the surrogate.  If we also preserve the locations of the {\it transitional}, {\it double}, and {\it single} labels, then we can compute new values for $d_\text{T-S}$ and $d_\text{D-S}$,
say  $\tilde{d}_\text{T-S}$ and $\tilde{d}_\text{D-S}$, from the surrogate. If we produce many surrogate sequences then we will get a sampling of  
surrogate values $\tilde{d}$, one for each of the two ambiguity indexes, to which we can compare $d$. We made several experiments of this type---for each of the two indexes, T-S and D-S, within each of the seven RNA families (Group I and Group II Introns, tmRNA, SRP RNA, RNase P, and 16s and 23s rRNA).

To make this precise, consider an RNA molecule with primary structure $p$ and comparative secondary structure $s$. Construct a segment ``histogram function,'' $\mathcal{H}(p)$, which is the 
vector of the number of times that each of the $4^4$ possible segments appears in $p$.\footnote{${\cal H}$ is made a vector by fixing an arbitrary ordering of all possible segments.}.
Let $\mathcal{P}(p)$ be the set of all permutations of the ordering of nucleotides in $p$, and let 
$\mathcal{E}(p)\subseteq\mathcal{P}(p)$ be the subset of permutations that preserve the frequencies of four-tuples:
\[
{\cal E}(p) = \{p'\in\mathcal{P}(p): {\cal H}(p')={\cal H}(p)\}
\]
Clever algorithms exist for efficiently drawing independent samples from the uniform distribution on $\mathcal{E}$---see \cite{Kandel1996-kz, Fitch1983-ec, Altschul1985-nq}. Let $p^{(1)},
\ldots,p^{(K)}$ be $K$ such samples, and let $\displaystyle
d_\text{T-S}\big(p^{(1)},s\big),\ldots,d_\text{T-S}\big(p^{(K)},s\big)$ and
$\displaystyle
d_\text{D-S}\big(p^{(1)},s\big),\ldots,d_\text{D-S}\big(p^{(K)},s\big)$ 
be the corresponding T-S and D-S ambiguity indexes. The results are not very sensitive to $K$, nor to the particular sample, provided that $K$ is large enough. We used $K=$10,000. Finally, let 
$\alpha_\text{T-S}(p,s) \in [0,1]$ and
$\alpha_\text{D-S}(p,s) \in [0,1]$
be the left-tail empirical probabilities, under the distributions defined by the ensembles
\begin{align}
&  d_\text{T-S}(p,s),d_\text{T-S}\big(p^{(1)},s\big),\ldots,d_\text{T-S}\big(p^{(K)},s\big) 
\label{eqn:T-Sensemble}\\
\text{and \hglue .8cm} 
& d_\text{D-S}(p,s),d_\text{D-S}\big(p^{(1)},s\big),\ldots,d_\text{D-S}\big(p^{(K)},s\big) 
\label{eqn:D-Sensemble}
\end{align}
of choosing an ambiguity index less than or equal to $d_\text{T-S}(p,s)$ and 
$d_\text{T-S}(p,s)$, respectively:
\begin{align}
\alpha_\text{T-S}(p,s)  & = \frac
{1 + \#\{k\in\{1,\ldots,K\}: d_\text{T-S}\big(p^{(k)},s\big) \leq d_\text{T-S}(p,s)\}}{1+K} \label{eqn:alpha_T-S} \\
\alpha_\text{D-S}(p,s)  & = \frac
{1 + \#\{k\in\{1,\ldots,K\}: d_\text{D-S}\big(p^{(k)},s\big)\leq d_\text{D-S}(p,s)\}}{1+K} \label{eqn:alpha_D-S}
\end{align}
In essence, each $\alpha$ score is a self-calibrated ambiguity index.

It is tempting to interpret $\alpha_\text{T-S}(p,s)$ as a p-value from a conditional hypothesis test: Given $s$ and $\mathcal{H}$, test the null hypothesis that $d_\text{T-S}(p,s)$ is statistically indistinguishable from  
$d_\text{T-S}(p^\prime,s)$, where $p^\prime$ is a random sample from $\mathcal{E}$. If the alternative hypothesis were that $d_\text{T-S}(p,s)$ is too small to be consistent with the null,  then the null is rejected in favor of the alternative with probability $\alpha_\text{T-S}(p,s)$. The problem with this interpretation, and the analogous one for $\alpha_\text{D-S}(p,s)$, is that this null hypothesis violates the simple observation that given $\mathcal{H}$ there is information in $s$ about $p$, whereas  $p^{(1)},\ldots,p^{(K)}$ are independent of $s$ given $\mathcal{H}$. A larger problem is that there is no reason to believe the alternative; we are more interested in {\em relative} than absolute ambiguity indexes. Thinking of  
$\alpha_\text{T-S}(p,s)$ and $\alpha_\text{D-S}(p,s)$ as calibrated intra-molecular indexes, we want to know how they vary across RNA families, and whether these variations depend on the differences between comparative and MFE structures.

@@

Does the MFE structure similarly separate single-entity RNA molecules from those that form ribonucleoproteins? A convenient way to explore this question is to recalculate and recalibrate the ambiguity indexes of each molecule in each of the seven families, but using the MFE in place of the comparative secondary structures. The results are summarized in Table \ref{table:explore_MFE}.
By comparison to the results shown from Table \ref{table:explore_comp}, 
the separation of unbound from bound molecules nearly disappears under the MFE secondary structures.
Possibly, the comparative structures, as opposed to the MFE structures, better anticipate the need to avoid energy traps in the folding landscape. Here too we will revisit the data using ROC curves and proper hypothesis tests.

@@

\subsection*{Formal Statistical Analyses}
The T-S and D-S ambiguity indexes ($d_\text{T-S}(p,s)$ and $d_\text{D-S}(p,s)$)  are intra-molecular measures of the extent to which segments within single-stranded regions have more Watson-Crick and wobble pairings than segments that lie in and around sub-sequences that are double-stranded. As such, they depend on both $p$ and any purported secondary structure, $s$. Based on calibrated versions of these indexes ($\alpha_\text{T-S}(p,s)$ and $\alpha_\text{D-S}(p,s)$) and employing the comparative secondary structure for $s$ we found support for the idea that non-coding RNA molecules which are active as individual entities are more likely to have small ambiguity indexes than RNA molecules destined to operate as part of ribonculeoproteins. Furthermore, the difference appears to be sensitive to the approach used for identifying secondary structure---there is little, if any, evidence in the MFE secondary structures for lower ambiguities among unbound molecules.

@@

As mentioned earlier, we can also associate a traditional p-value to the problem of separating unbound from bound molecules, based on the T-S ambiguity indexes. We consider only the signs (positive or negative) of these indexes, and then test whether there are fewer than expected positive indexes among the unbound, as opposed to the bound, population. This amounts to computing $\mathbb{P}\{N\leq n_1\}$ from the hypergeometric distribution---Eq (\ref{eqn:null}). The relevant statistics can be found in Table \ref{table:NumberPositive}, 
under the column labels {\bf \#mol's} and $\bm{\#d_\text{\bf T-S}>0}$. Specifically, 
$M_1=116+34=150$ (number of unbound molecules),
$M_2=404+346+407+279+59=1495$ (number of bound molecules), $n_1=50+8=58$ (number of positive T-S indexes among unbound molecules) and $n_2=368+269+379+210+53=1279$.
The resulting p-value, $1.2\cdot 10^{-34}$, is essentially zero, meaning that the positive T-S indexes are not distributed proportional to the sizes of the unbound and bound populations, which is by now obvious, in any case. To repeat our caution, small p-values conflate sample size with effect size, and for that reason we have chosen additional ways, using permutations as well as classifications, to look at the data.

@@

\section*{Discussion}
We have introduced two indexes designed as coarse measures of the prevalence of kinetic traps in the energy landscape of a structural RNA molecule. These indexes, which we call the T-S and D-S ambiguity indexes, are each a function of both the primary and any presumed secondary structure of the molecule. The first, the T-S index, compares the availability of false pairings (i.e. stems that are not part of the purported secondary structure) of sequences straddling transitions from double to single-stranded regions to the availability of false pairings strictly within single-stranded regions.
The second, the D-S index, results from a similar calculation, but using double-stranded sequences in place of {\em transitional} sequences. We have given evidence that either index can be used effectively to classify unbound RNA (molecules that do not act as components of ribonucleoproteins) from bound RNA (those that are components of ribonucleoproteins), {\em provided that the secondary structure is derived from comparative analysis}. 
Specifically, unbound molecules typically have lower ambiguity indexes than bound molecules. On the other hand, there is little evidence for this effect in the ambiguity indexes derived instead from MFE analyses, and classifiers based on MFE secondary structure are substantially less accurate. Furthermore, either index can be used in unbound molecules to effectively classify a given secondary structure as derived from comparative as opposed to MFE analysis, but not in bound molecules. In particular, among unbound molecules comparative-analysis structures typically have lower ambiguity indexes than MFE structures.

@@

Among our goals is to stress the possible importance of kinetics, as opposed to thermal equilibrium {\em per se}, in the determination of native structure. Demonstrably, many non-coding RNA molecules will spontaneously fold and then remain in apparent equilibrium in their biologically active structures. 
But biological stability is relative to biological timescale; the folding of any particular RNA could just as well end in metastability, provided that the process is repeatable and the result sufficiently stable over the molecule's proper biological lifetime. Indeed, it would be arguably easier to evolve an effective tertiary structure without the additional and unnecessary burden of thermal equilibrium. 
However, and this has often been pointed out (e.g. \cite{Ding2003-sv, Mathews2006-th}), the MFE structure itself may be a poor representative of thermal equilibrium. It is quite possible, then, that our observations to the effect that comparative and MFE structures have substantially different relationships to the ambiguity indexes, and our interpretation that comparative structures better separate unbound from bound molecules, would not hold up if we were to adopt a more ensemble-oriented structure in place of the MFE, as advocated by \cite{Ding2005-br}, for example.

Our expectation was that {\em transitional} and {\em double} locations would be relatively more protected from kinetic traps than {\em single} locations, and hence the focus on D and T relative to S in the formulation of the indexes. But this coin has another side: high ambiguity of {\em single} locations could certainly discourage, if not actually stereochemically prevent, access to native stems. In a related vein, Lin et al. \cite{Lin2018-xl} have given evidence that competing stems which are inconsistent may {\em both} contain a high measure of information about the equilibrium distribution, suggesting that in such cases both forms could be active and the notion of {\em single} might itself be ambiguous. This analysis, though, does presume the relevance of the equilibrium distribution, against which Levinthal has made some compelling arguments.

@@

\subsection*{RNA Secondary Structure Prediction Methods}
Comparative analysis\cite{Gutell1992-hu} is based on the simple principle that a single RNA secondary structure can be formed from different RNA sequences. It works with alignments of homologous sequences, and is highly accurate in determining RNA secondary structures\cite{Gutell2002-ei}. In this paper, we used a large set of RNA secondary structures determined by carefully conducted comparative analysis  to serve as the ground truth.

>>>>

% Please keep the abstract below 300 words
\section*{Abstract}
The route to folding for a large non-coding RNA molecule is full of 
kinetic traps---energetically favorable pairings of nucleotide sequences that are not part of the active structure. Yet many of these molecules spontaneously fold to a native structure when placed in a suitable chemical environment at an appropriate temperature.
Levinthal's convincing arguments\cite{Levinthal1969-hi} (aka Levinthal's paradox) about proteins apply to RNA molecules as well, perhaps even more compellingly: arrival to thermal equilibrium should take a great deal more time than the biological lifetime of the organism. Evidently, mechanisms have evolved that contribute to a directed and efficient folding process, and quite possibly the active configuration is not at all like a sample from an equilibrium distribution, but rather a metastable state, provided only that it will typically last long enough for its purpose. 
Based on the number of possible Watson-Crick or G$\cdot$U wobble pairings at each location, we
introduce various measures of the difficulty in reaching a stable configuration, whether or not it is a true equilibrium.
We call these measures ``ambiguity indexes,'' and focus on two, each of which is a function of both the primary and any given proposed secondary structure. Comparing families of RNA molecules that appear in ribonucleoproteins (``bound'' molecules) to RNA molecules that do not operate as nucleoproteins (``unbound'' molecules), we find that the unbound molecules have systematically lower ambiguity indexes, {\em provided that the indexes are computed using the comparative-analysis \cite{Gutell1992-hu} structure} (also known as the ``gold standard''). The effect largely disappears when the indexes are based, instead, on the so-called minimum-free energy (MFE) structures \cite{Mathews1999-hc,Zuker1999-rc}. Since MFE structures are derived from an approximation of the equilibrium distribution, we interpret these and other related results as evidence for metastable but not necessarily thermal equilibrium secondary structure.

@@

Because of the time-consuming nature of experimental determination of RNA structures, a considerable amount of work has been put into computational (as opposed to experimental) approaches. For secondary structure prediction, comparative analysis is the gold standard  \cite{Gutell1992-hu}. But accurate comparative analysis typically requires a large number of homologous sequences, which may or may not be available. On the other hand, the prevailing {\em computational} approaches are based on an important and controversial assumption\cite{Flamm2008-ir}. Namely, that the structures of  non-coding RNA molecules, {\em in vivo}, are in thermal equilibrium. If this were the case, and if the configuration energies could be accurately specified, then secondary (and in principle, even tertiary) structures could be explored through Monte Carlo sampling from the Boltzmann equilibrium distribution. Alternatively, depending on the nature of the energy landscape, secondary structures could be approximated by identifying the minimum free energy (MFE) configuration  \cite{Mathews1999-hc,Zuker1999-rc}. However, even if we put aside the necessity for approximations, the biological relevance of equilibrium configurations has been a source of misgivings at least since 1969, when Levinthal pointed out that the time required to equilibrate might be too long by many orders of magnitude  \cite{Levinthal1969-hi}. In light of these observations, and considering the ``frustrated'' nature of the folding landscape, many have argued that when it comes to structure prediction for macromolecules, kinetic accessibility is more relevant than equilibrium thermodynamics  \cite{Higgs2000-gj,Flamm2008-ir,Baker1994-px}. In fact, a metastable state that is sufficiently long-lived and accessible might be biologically indistinguishable from an equilibrium state.

The primary structures of RNA molecules typically afford many opportunities to form short or medium-length stems,\footnote{
By which we will mean sequences of G$\cdot$U (``wobble pairs'') and/or Watson-Crick pairs.}
most of which do not participate in the native structure. This situation not only makes it hard for the biologist to accurately predict secondary structure, but equally challenges the molecule to avoid these ``energetic traps.'' Once formed, they require a large amount of energy (not to mention time) to be unformed. By what mechanisms might these traps be discouraged, if not actually avoided? 
Ideally, the only stems available in the primary structure would be those that participate in the secondary structure, but this is obviously too restrictive for all but the shortest or simplest of the structural RNA molecules. Still, there can be little doubt that the necessity for repeatable and efficient folding produces a selective pressure against the more disruptive ambiguities---e.g.
a sequence that can form half of either of two stems, both of which are kinetically accessible and stable, but only one of which is native, or a sequence that is unpaired in the native structure but can nonetheless form a stem that is stereochemically inconsistent with the native structure. By this reasoning, which emphasizes kinetics rather than equilibrium, {\em per se}, we might expect to find information in the intra-molecular ambiguities about the folding process, or even the native structure.

We will introduce quantitative measures of ambiguity and demonstrate their statistical relationships to native secondary structure and to qualitative distinctions between RNA families.
Using a somewhat arbitrary convention, we will
refer to sequences of length four (four consecutive nucleotides) as {\em segments}. Keeping this convention in mind, each line of statistical
evidence is based on the same formal definition of the ``local ambiguity'' of a given segment. This is simply the number of its complementary pairs in the molecule. 
When we refer to the {\em location} of a segment, we will mean the location of the first element of the segment, counting from 5$^\prime$ to 3$^\prime$, and  
when we refer to the local ambiguity of a location, we will mean the local ambiguity of the segment at that location. 
Local ambiguity is an intrinsic property of the primary structure. We will be interested in exposing its statistical relationships to any given candidate secondary structure of the same molecule. For this purpose, for any particular secondary structure we distinguish three different kinds of locations:
\begin{description}
	\item[\textit{Single}:] Locations where all nucleotides in the corresponding segment are unpaired in the secondary structure;
	\item[\textit{Double}:] Locations where all nucleotides in the corresponding segment are paired in the secondary structure;
	\item[\textit{Transitional}:] Locations where some nucleotides in the corresponding segment are paired and others are unpaired in the secondary structure.
\end{description}
{\em Double} and {\em transitional} locations participate in the candidate secondary structure, while {\em single} locations do not.
As a general rule, local ambiguities measured at any of these locations will increase with the length of the molecule---there are simply more opportunities to find complementary sequences.
Instead of using the local ambiguities themselves, we focus on the {\em differences} between ambiguities in and around stems ({\em double} and {\em transitional} locations) from those at unpaired ({\em single}) locations. In particular, we defined the ``T-S ambiguity index'' to be the difference between the average local ambiguity at {\em transitional} locations minus the average at {\em single} locations. We defined the ``D-S ambiguity index'' similarly, but adjusted for an evident bias at {\em double} locations, most of which will have at least one complementary pair somewhere in the molecule. 
(The exceptions come from bulges, which account for about 20\% of the {\em double} locations in our data set.
By our definition, segments are {\em consecutive} sequences. A segment that is part of a bulge does not necessarily match any consecutive sequence of four nucleotides.)
With this in mind, we defined the D-S ambiguity index to be the average local ambiguity at those {\em double} locations that have at least one ambiguity minus the average over those {\em single} locations that have at least one ambiguity.\footnote{
An argument could be made for including bulges in the definition of the {\em local ambiguity} in the first place. We have not explored this or other variations, though there is not shortage of arguably useful descriptive statistics for capturing the notion of ambiguity.}

@@

How are
we to gauge $d_\text{T-S}$ and $d_\text{D-S}$, and how are we to compare their values across different RNA families? 
Consider the following experiment: for a given RNA molecule we create a ``surrogate'' which has the same nucleotides, and in fact the same counts of {\em all} four-tuple segments as the original molecule, but is otherwise ordered randomly. If ACCU appeared eight times in the original molecule, then it appears eight times in the surrogate, and the same can be said of all sequences of four successive nucleotides---the frequency of each of the $4^4$ possible segments is preserved in the surrogate.  If we also preserve the locations of the {\it transitional}, {\it double}, and {\it single} labels, then we can compute new values for $d_\text{T-S}$ and $d_\text{D-S}$,
say  $\tilde{d}_\text{T-S}$ and $\tilde{d}_\text{D-S}$, from the surrogate. If we produce many surrogate sequences then we will get a sampling of  
surrogate values, $\tilde{d}_\text{T-S}$ and $\tilde{d}_\text{D-S}$, to which we can compare $d_\text{T-S}$ and $d_\text{D-S}$, respectively. We made several experiments of this type---for each of the two indexes, T-S and D-S, within each of the seven RNA families (Group I and Group II Introns, tmRNA, SRP RNA, RNase P, and 16s and 23s rRNA).

To make this precise, consider an RNA molecule with primary structure $p$ and comparative secondary structure $s$. Construct a segment ``histogram function,'' $\mathcal{H}(p)$, which is the 
vector of the number of times that each of the $4^4$ possible segments appears in $p$.\footnote{${\cal H}$ is made a vector by fixing an arbitrary ordering of all possible segments.}
Let $\mathcal{P}(p)$ be the set of all permutations of the ordering of nucleotides in $p$, and let 
$\mathcal{E}(p)\subseteq\mathcal{P}(p)$ be the subset of permutations that preserve the frequencies of four-tuples:
\[
{\cal E}(p) = \{p'\in\mathcal{P}(p): {\cal H}(p')={\cal H}(p)\}
\]
Clever algorithms exist for efficiently drawing independent samples from the uniform distribution on $\mathcal{E}$---see \cite{Kandel1996-kz, Fitch1983-ec, Altschul1985-nq}. Let $p^{(1)},
\ldots,p^{(K)}$ be $K$ such samples, and let $\displaystyle
d_\text{T-S}\big(p^{(1)},s\big),\ldots,d_\text{T-S}\big(p^{(K)},s\big)$ and
$\displaystyle
d_\text{D-S}\big(p^{(1)},s\big),\ldots,d_\text{D-S}\big(p^{(K)},s\big)$ 
be the corresponding T-S and D-S ambiguity indexes. The results are not very sensitive to $K$, nor to the particular sample, provided that $K$ is large enough. We used $K=$10,000. Finally, let 
$\alpha_\text{T-S}(p,s) \in [0,1]$ and
$\alpha_\text{D-S}(p,s) \in [0,1]$
be the left-tail empirical probabilities, under the distributions defined by the ensembles
\begin{align}
&  d_\text{T-S}(p,s),d_\text{T-S}\big(p^{(1)},s\big),\ldots,d_\text{T-S}\big(p^{(K)},s\big) 
\label{eqn:T-Sensemble}\\
\text{and \hglue .8cm} 
& d_\text{D-S}(p,s),d_\text{D-S}\big(p^{(1)},s\big),\ldots,d_\text{D-S}\big(p^{(K)},s\big) 
\label{eqn:D-Sensemble}
\end{align}
of choosing an ambiguity index less than or equal to $d_\text{T-S}(p,s)$ and 
$d_\text{T-S}(p,s)$, respectively:
\begin{align}
\alpha_\text{T-S}(p,s)  & = \frac
{1 + \#\{k\in\{1,\ldots,K\}: d_\text{T-S}\big(p^{(k)},s\big) \leq d_\text{T-S}(p,s)\}}{1+K} \label{eqn:alpha_T-S} \\
\alpha_\text{D-S}(p,s)  & = \frac
{1 + \#\{k\in\{1,\ldots,K\}: d_\text{D-S}\big(p^{(k)},s\big)\leq d_\text{D-S}(p,s)\}}{1+K} \label{eqn:alpha_D-S}
\end{align}
In essence, each $\alpha$ score is a self-calibrated ambiguity index.

It is tempting to interpret $\alpha_\text{T-S}(p,s)$ as a p-value from a conditional hypothesis test: Given $s$ and $\mathcal{H}$, test the null hypothesis that $d_\text{T-S}(p,s)$ is statistically indistinguishable from  
$d_\text{T-S}(p^\prime,s)$, where $p^\prime$ is a random sample from $\mathcal{E}$. If the alternative hypothesis were that $d_\text{T-S}(p,s)$ is too small to be consistent with the null,  then the null is rejected in favor of the alternative with probability $\alpha_\text{T-S}(p,s)$. The problem with this interpretation, and the analogous one for $\alpha_\text{D-S}(p,s)$, is that this null hypothesis violates the observation that given $\mathcal{H}$ there is information in $s$ about $p$, whereas  $p^{(1)},\ldots,p^{(K)}$ are independent of $s$ given $\mathcal{H}$. 
In other words, $d_\text{T-S}(p,s)$ and $d_\text{T-S}(p^\prime,s)$
have different conditional distributions given $s$ and $\mathcal{H}$, in direct contradiction to the null hypothesis.
A larger problem is that there is no reason to believe the alternative; we are more interested in {\em relative} than absolute ambiguity indexes. Thinking of  
$\alpha_\text{T-S}(p,s)$ and $\alpha_\text{D-S}(p,s)$ as calibrated intra-molecular indexes, we want to know how they vary across RNA families, and whether these variations depend on the differences between comparative and MFE structures.

@@

Does the MFE structure similarly separate single-entity RNA molecules from those that form ribonucleoproteins? A convenient way to explore this question is to recalculate and recalibrate the ambiguity indexes of each molecule in each of the seven families, but using the MFE in place of the comparative secondary structures. The results are summarized in Table \ref{table:explore_MFE}.
By comparison to the results shown from Table \ref{table:explore_comp}, 
the separation of unbound from bound molecules nearly disappears under the MFE secondary structures.
Possibly, the comparative structures, as opposed to the MFE structures, better anticipate the need to avoid energy traps in the folding landscape. Here too we will soon revisit the data using ROC curves and proper hypothesis tests.

@@

\subsection*{Formal Statistical Analyses}
The T-S and D-S ambiguity indexes ($d_\text{T-S}(p,s)$ and $d_\text{D-S}(p,s)$)  are intra-molecular measures of the extent to which segments that lie in and around sub-sequences that are double-stranded have more Watson-Crick and wobble pairings than
segments within single-stranded regions. As such, they depend on both $p$ and any purported secondary structure, $s$. Based on calibrated versions of these indexes ($\alpha_\text{T-S}(p,s)$ and $\alpha_\text{D-S}(p,s)$) and employing the comparative secondary structure for $s$ we found support for the idea that non-coding RNA molecules which are active as individual entities are more likely to have small ambiguity indexes than RNA molecules destined to operate as part of ribonculeoproteins. Furthermore, the difference appears to be sensitive to the approach used for identifying secondary structure---there is little, if any, evidence in the MFE secondary structures for lower ambiguities among unbound molecules.

@@

As mentioned earlier, we can also associate a traditional p-value to the problem of separating unbound from bound molecules, based on the T-S ambiguity indexes. We consider only the signs (positive or negative) of these indexes, and then test whether there are fewer than expected positive indexes among the unbound, as opposed to the bound, population. This amounts to computing $\mathbb{P}\{N\leq n_1\}$ from the hypergeometric distribution---Eq (\ref{eqn:null}). The relevant statistics can be found in Table \ref{table:NumberPositive}, 
under the column labels {\bf \#mol's} and $\bm{\#d_\text{\bf T-S}>0}$. Specifically, 
$M_1=116+34=150$ (number of unbound molecules),
$M_2=404+346+407+279+59=1495$ (number of bound molecules), $n_1=50+8=58$ (number of positive T-S indexes among unbound molecules) and $n_2=368+269+379+210+53=1279$.
The resulting p-value, $1.2\cdot 10^{-34}$, is essentially zero, meaning that the positive T-S indexes are not distributed proportional to the sizes of the unbound and bound populations, which is by now obvious in any case. To repeat our caution, small p-values conflate sample size with effect size, and for that reason we have chosen additional ways, using permutations as well as classifications, to look at the data.

@@

\section*{Discussion}
We have introduced two indexes designed as coarse measures of the prevalence of kinetic traps in the energy landscape of a structural RNA molecule. These indexes, which we call the T-S and D-S ambiguity indexes, are each a function of both the primary and any presumed secondary structure of the molecule. The first, the T-S index, compares the availability of pairings (i.e. stems that are not part of the purported secondary structure) of sequences straddling transitions from double to single-stranded regions to the availability of pairings strictly within single-stranded regions.
The second, the D-S index, results from a similar calculation, but using double-stranded sequences in place of {\em transitional} sequences. We have given evidence that either index can be used effectively to classify unbound RNA (molecules that do not act as components of ribonucleoproteins) from bound RNA (those that are components of ribonucleoproteins), {\em provided that the secondary structure is derived from comparative analysis}. 
Specifically, unbound molecules typically have lower ambiguity indexes than bound molecules. On the other hand, there is little evidence for this effect in the ambiguity indexes derived instead from MFE analyses, and classifiers based on MFE secondary structure are substantially less accurate. Furthermore, either index, T-S or D-S, can be used in unbound molecules to effectively classify a given secondary structure as derived from comparative as opposed to MFE analysis, but not in bound molecules. In particular, among unbound molecules comparative-analysis structures typically have lower ambiguity indexes than MFE structures.

@@

Among our goals is to stress the possible importance of kinetics, as opposed to thermal equilibrium {\em per se}, in the determination of native structure. Demonstrably, many non-coding RNA molecules will spontaneously fold and then remain in apparent equilibrium in their biologically active structures. 
But biological stability is relative to biological timescale; the folding of any particular RNA could just as well end in metastability, provided that the process is repeatable and the result sufficiently stable over the molecule's proper biological lifetime. Indeed, it would be arguably easier to evolve an effective tertiary structure without the additional and unnecessary burden of thermal equilibrium. 
However, and this has often been pointed out (e.g. \cite{Ding2003-sv, Mathews2006-th}), the MFE structure itself may be a poor representative of thermal equilibrium. It is  possible, then, that our observations to the effect that comparative and MFE structures have substantially different relationships to the ambiguity indexes, and our interpretation that comparative structures better separate unbound from bound molecules, would not hold up as well if we were to adopt a more ensemble-oriented structure in place of the MFE, as advocated by \cite{Ding2005-br}, for example.

Our expectation was that {\em transitional} and {\em double} locations would be relatively more protected from kinetic traps than {\em single} locations, and hence the focus on D and T relative to S in the formulation of the indexes. But this coin has another side: high ambiguity of {\em single} locations could certainly discourage, if not actually stereochemically prevent, access to native stems. In a related vein, Lin et al. \cite{Lin2018-xl} have given evidence that competing stems which are inconsistent may {\em both} contain a high measure of information about the equilibrium distribution, suggesting that in such cases both forms could be active and the notion of {\em single} might itself be ambiguous. This analysis, though, does presume the relevance of the equilibrium distribution, against which Levinthal's observations make for some compelling arguments.

@@

\subsection*{RNA Secondary Structure Prediction Methods}
Comparative analysis\cite{Gutell1992-hu} is based on the simple principle that a single RNA secondary structure can be formed from different RNA sequences. Using alignments of homologous sequences, comparative analysis has proven to be highly accurate in determining RNA secondary structures\cite{Gutell2002-ei}. We used a large set of RNA secondary structures determined by comparative analyses to serve as ground truth.
